Narrative

### 2023-05-20 manuscript/introduction.qmd
intro: dropped last paragraph of no-one talking about it

Imperfect reliability in primary studies imbues heterogeneity estimates with both a positive and a negative bias. The positive bias is discussed by several highly cited meta-analytic handbooks, most prominently by Hunter and Schmidt (2015; 12,235 citations on google scholar as of 2023-05-11) who introduced the concept of psychometric meta-analysis (i.e., meta-analysis corrected for measurement artifacts such as unreliability), but also by Borenstein et al. (2009, p. 342; 19,020 citations), Card (2015, p.126; 1,716 citations) as well as in recent literature (Wiernick and Dahlke, 2020). However, there appears to be a lack of acknowledgement that imperfect reliability can also lead to a negative bias in heterogeneity estimates. The only one of the above sources to refer to a negative bias is Hunter and Schmidt who mention it shortly in their 2nd edition of *Methods of Meta-analysis* (2004, p. 142) in a paragraph later absent from the 3rd edition (Hunter & Schmidt, 2015). Below we detail the argument for both a positive and negative bias in heterogeneity estimates.

@@

In classical test theory (CTT), reliability of measurement ($R_{xx}$) is defined as the proportion true score variance ($\sigma^2_T$) to observed score variance ($\sigma^2_X$) and is presumed to be constant for all participants in a sample $R_{xx} = \sigma^2_T / \sigma^2_X$. We use unreliability of measurement to refer to $1 - R_{xx}$. Imperfect reliability in measurements leads to attenuation of observed effect sizes. For product-moment correlations the observed correlation can be computed as $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ where $r_{xy}$ is the observed correlation between the variables $X$ and $Y$, $\rho_{xy}$ is the true correlation (with or without sampling error) and $R_{xx'}$ and $R_{yy'}$ are the measurement reliabilities for $X$ and $Y$. As such, correcting for unreliability is straightforward for product-moment correlations, but can also be done for other effect size types (Wiernick and Dahlke, 2020). A corrected correlation should also have its sampling variance corrected. This is inflated by a factor equal to the square of the reliabilities, such that if $a = \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ then the corrected sampling variance $V_\rho = V_{r} / a^2$ (e.g., Borenstein, 2009, p. 343).

>>>>

@@

In classical test theory (CTT), reliability of measurement ($R_{xx}$) is defined as the proportion true score variance ($\sigma^2_T$) to observed score variance ($\sigma^2_X$) and is presumed to be constant for all participants in a sample $R_{xx} = \sigma^2_T / \sigma^2_X$. We use unreliability of measurement to refer to $1 - R_{xx}$. Imperfect reliability in measurements leads to attenuation of observed effect sizes. For product-moment correlations the observed correlation can be computed as $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ where $r_{xy}$ is the observed correlation between the variables $X$ and $Y$, $\rho_{xy}$ is the true correlation (with or without sampling error) and $R_{xx'}$ and $R_{yy'}$ are the measurement reliabilities for $X$ and $Y$. As such, correcting for unreliability is straightforward for product-moment correlations, but can also be done for other effect size types (Wiernick and Dahlke, 2020). A corrected correlation should also have its sampling variance corrected. This is inflated by a factor equal to the square of the reliabilities, such that if $a = \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ then the corrected sampling variance $V_\rho = V_{r} / a^2$ (e.g., Borenstein, 2009, p. 343). Imperfect reliability in primary studies imbues heterogeneity estimates with both a positive and a negative bias.
